iSEE: Interactive SummarizedExperiment Explorer

### 2018-03-30 manuscript/main.tex
Incorporated Figure 1, removed subheadings to save space.

Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits the widely-used SummarizedExperiment data architecture from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
This enables iSEE to accommodate a wide range of `-omics datasets that are routinely processed by Bioconductor packages, ranging from RNA sequencing to single-cell transcriptomics to mass cytometry.

\section{Description of the iSEE interface}

\subsection{Integration with Bioconductor packages} 
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results. 
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
Developers can easily derive subclasses for specialized use cases, such as the SingleCellExperiment class that stores sample-wise reduced dimensionality embeddings for single-cell `-omics data analyses. 
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in the R/Bioconductor framework. 
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

\subsection{Visualizing multiple data aspects simultaneously}
iSEE uses a customisable multi-panel layout that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
This includes metadata for each feature or sample, experimental observations across samples for each feature, and reduced dimensionality embeddings for each sample. 
For most panels, each sample or feature is represented as a point in a plot with the specified variables on the x/y-axes.
Depending on whether these variables are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the number of samples/features with that combination). 
Each point can also be coloured according to data aspects such as sample-level metadata or feature-level observations. 
Other panel types include heatmaps, to visualize experimental observations for multiple features; and row statistics tables, to facilitate searching across features and their metadata.
All panels can be removed, reordered or resized by the user, and new panels of any type can be added to the interface.

\subsection{Dynamic linking across panels}
A key feature of iSEE is the ability to dynamically transmit information between panels. 
Users can define arbitrary links between a ``transmitting'' and ``receiving'' plot, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot (or optionally, hide all unselected points). 
This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to a feature expression plot. 
The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object. 
Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.

\subsection{Customizing the user experience}
iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each individual  metadata variable and assay type consistently across all panels in the application. 
Row statistics tables can be augmented with dynamic annotation based on the selected row, providing links to online resources such as Ensembl or Entrez. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

@@

\section*{Acknowledgements}
We thank the organizers of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions. 
\vspace*{-12pt}

>>>>

\begin{figure*}[t]
\includegraphics[width=\textwidth]{pics/Fig1.pdf}
\caption{iSEE uses a customisable multi-panel layout (A) that simultaneously displays one or more panels of various types, where each panel type visualizes a different aspect of the data. 
New panels of any type can be added (i), and all panels can be removed, reordered or resized (ii).
Panel types are available visualize sample-based reduced dimensionality embeddings (iii), sample-level metadata (iv), and experimental observations across samples for each feature (v). 
For these panel types, each sample is represented as a point, which can be coloured according to sample-level metadata or feature-level observations. 
Depending on whether the variables of interest are categorical or numeric, panels will automatically switch between scatter plots, violin plots or ``rectangle plots'' (where each combination of categorical levels is represented by a rectangle with area proportional to the frequency of combination). 
Other panel types include row statistics tables (vi), to facilitate searching across features and their metadata; heatmaps (vii), to visualize experimental observations for multiple features; and feature-level metadata plots.
(B) Information can be transmitted between panels according to a user-specified scheme.
Here, the selection of feature $X$ in the row statistics table determines the y-axis of the feature expression plot, and colours the samples in the reduced dimension plot by the expression of $X$.
Selection of points in the reduced dimension plot (dotted blue line) also determines the samples that are shown in the column data (i.e., sample metadata) plot; 
further selection of points in the column data plot determines the samples that are shown in the heatmap.
}
\label{fig:iSEE}
\end{figure*}

\section{Description of the iSEE interface}
Here, we present the iSEE software package for interactive data exploration.
iSEE is implemented in R using the shiny framework \citep{chang2017shiny} and exploits data architectures from the open-source Bioconductor project \citep{gentleman2004bioconductor}.
The iSEE interface is initialized with a single function call accepting a SummarizedExperiment object as input \citep{huber2015orchestrating}.
This object stores one or more matrices of experimental observations as ``assays'', where columns and rows represent samples and features, respectively. 
Per-feature or per-sample variables are stored in the ``rowData'' and ``colData'', respectively; this includes experimental metadata as well as any analysis results.
All of these data aspects can be directly examined in the multi-panel layout of the iSEE interface (Fig.~\ref{fig:iSEE}A).
The flexibility of the SummarizedExperiment class is the driving factor behind its broad deployment throughout the Bioconductor ecosystem, with uses in RNA sequencing \citep{love2014moderated}, methylation \citep{aryee2014minfi} and Hi-C analysis pipelines \citep{lun2016infrastructure}, amongst others.
By accepting a SummarizedExperiment object, iSEE immediately offers interactive visualization for a variety of data modalities, complementing the state-of-the-art analysis methodologies already available in R/Bioconductor packages. 
%In particular, we leverage off the state-of-the-art allowing us to avoid re-implementing them in iSEE (as is commonly done in other standalone visualization tools).

A key feature of iSEE is the ability to dynamically transmit information between panels (Fig.~\ref{fig:iSEE}B).
Users can define arbitrary links between ``transmitting'' and ``receiving'' plots, whereby selection of points on the transmitting plot will highlight the corresponding points in the receiving plot. 
This facilitates exploration of the relationships between different aspects of the data -- for example, users can easily determine co-expression patterns of genes in a particular region of a reduced dimensionality embedding, by transmitting information from the reduced dimension plot to a feature expression plot. 
The linking paradigm extends to multiple plots whereby a plot can transmit to multiple receivers, or a receiving plot can itself transmit to another plot.
This allows users to mimic the arbitrarily complex gating strategies often found in analyses of flow cytometry data \citep{finak2014opencyto}, with the advantage of being extensible to any (meta)data present in a SummarizedExperiment object. 
Information can similarly be transferred between tables and plots, e.g., selecting a row determines how points are coloured in receiving plots.

iSEE also offers a host of other customizable features to improve data exploration. 
Users can supply their own colour mappings to iSEE, to control how colour scales are chosen for each metadata variable and assay type across all panels in the application.
Row statistics tables can be augmented with dynamic annotation based on the selected row, linking to online resources such as Ensembl or Entrez. 
For large datasets, points can be downsampled in a density-dependent manner to accelerate rendering of the plots, improving the responsiveness of the interface without compromising the fidelity of the visualization. 
After analyses are completed, users can deploy iSEE on a server with a bespoke step-by-step ``tour'' of their dataset, guiding the audience through an examination of the salient features in the data.
iSEE also memorizes the exact R code that was used to generate every plot. 
This allows users to easily reproduce the results of any exploratory analysis, simply by copying the code reported by iSEE into their own analysis scripts.

@@

\section*{Acknowledgements}
We thank the organizers and participants of the European Bioconductor Meeting 2017, where the idea for this package was first conceived.
We also thank members of the Bioconductor community for their helpful suggestions. 
\vspace*{-12pt}
